Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-29 survey/survey.tex
a bit more on data I

\newtheorem{defn}{Definition}[section]
\newtheorem{prop}{Proposition}[section]

@@

What does it mean for two finite sets to be ``the same''?  Well,
clearly the sets $A = \left\{1, 2, 3\right\}$ and
$B = \left\{c, d\right\}$ are different.  Why?  Well, supposed there
was a transformation $f : A \rightarrow B$ that deformed $A$ into $B$,
and another $g : B \rightarrow A$ which undid this
transformation. Since $f$ is total, by the pigeonhole principle, two
elements of $A$ would be mapped to the same element of $B$. Supposed
that this is $2$ and $3$, and that they both map to $d$.  But $g(d)$
cannot be both $2$ and $3$, and so $g$ is not the inverse of $f$. With
just a little more work, we can show that $f$ (and $g$) must be both
injective and surjective. In other words, $f$ (and $g$) must be a
bijection between $A$ and $B$.  And of course this only happens when
$A$ and $B$ have the same number of elements. More importantly, given
a bijection $f : C \rightarrow D$ of finite sets $C,D$, there always
exists another bijection $g : D \rightarrow C$ which is $f$'s
inverse. So, for finite sets, \emph{bijections} act as reversible
deformations.

@@

Bijections on \fin{n} have a specific name: permutations. As is well-known
permutations can be generated by sequences of transpositions. Thus we
can create a small language for writing permutations as
\[\begin{array}{rcl}
p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p;p
  \end{array}\]
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
$\mathit{swap}~j~j$ would represent the identity permutation.

While we are mainly interested in typed data, it is always worthwhile
to first start investigating the untyped setting first, as it gives us
insights into what the \emph{purely operational} view of the theory
would be.  This tells us that permutations are an inescapable part of
the fabric of reversible computing.  However as permutations are
untyped, and act on the canonicalized version of $n$-element sets
(i.e. those sets where all the structure has been forgotten), these
are a rather pale shadow of the rich tapestry of
information-perserving transformation of structured data which we
investigate next.

>>>>

\newtheorem{thm}{Theorem}[section]
\newtheorem{defn}{Definition}[section]
\newtheorem{prop}{Proposition}[section]

@@

What does it mean for two finite sets to be ``the same''?  Well,
clearly the sets $A = \left\{1, 2, 3\right\}$ and
$B = \left\{c, d\right\}$ are different.  Why?  Well, supposed there
was a transformation $f : A \rightarrow B$ that deformed $A$ into $B$,
and another $g : B \rightarrow A$ which undid this
transformation. Since $f$ is total, by the pigeonhole principle, two
elements of $A$ would be mapped to the same element of $B$. Suppose
that this is $2$ and $3$, and that they both map to $d$.  But $g(d)$
cannot be both $2$ and $3$, and so $g$ is not the inverse of $f$. With
just a little more work, we can show that $f$ (and $g$) must be both
injective and surjective. In other words, $f$ (and $g$) must be a
bijection between $A$ and $B$.  And of course this only happens when
$A$ and $B$ have the same number of elements. More importantly, given
a bijection $f : C \rightarrow D$ of finite sets $C,D$, there always
exists another bijection $g : D \rightarrow C$ which is $f$'s
inverse. So, for finite sets, \emph{bijections} act as reversible
deformations.

@@

Bijections on \fin{n} have a specific name: permutations. As is well-known
permutations can be generated by sequences of transpositions. Thus we
can create a small language for writing permutations as
\[\begin{array}{rcl}
p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p \fatsemi p
  \end{array}\]
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
$\mathit{swap}~j~j$ would represent the identity permutation.

Although this language appears weak, it is universal for boolean
combinational circuits. The result follows from the following
construction of the Toffoli gate which is universal for combinational
circuits. We write $[2^n]$ for the finite set representing $n$-bit
words with the canonical ordering for binary numbers. Thus when $n=3$,
the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which correspond to
the 3-bit words $\{000,001,010,011,100,101,110,111\}$. The Toffoli
gate is a reversible function $[2^3] \rightarrow [2^3]$ which swaps
$110$ and $111$, i.e., it can be realized as $\mathit{swap}~6~7$.

In programming practice we are however more interested in structured
data which will be the subject on next section. What we learn from
this short investigation using untyped flat sets is what the
\emph{purely operational} view of the theory would be. In particular,
it tells us that permutations are an inescapable part of the fabric of
reversible computing.  However as permutations are untyped, and act on
the canonicalized version of $n$-element sets (i.e. those sets where
all the structure has been forgotten), these are a rather pale shadow
of the rich tapestry of information-perserving transformation of
structured data which we investigate next.
